# Week 5: Analysis of longitudinal data

```{r}
date()
```

<br /> 

Note: As you read the data sets in your chapter6.Rmd you may have to factor the categorical variables again, as R may not recognise them automatically as factors.

Note that you must SWAP the data sets! :) It is NOT a simple copy & paste of my book!

## RATS (7p)

<br /> *Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)*

```{r,fig.dim= c(8,6)}
library(dplyr)
library(ggplot2)
library(GGally)
library(tidyr)
#library(boot)
library(corrplot)
library(FactoMineR)

# load the data
human <- read.csv("data/human.csv", header = TRUE, stringsAsFactors = TRUE, row.names = 1)

# summaries of the variables
summary(human)

# graphical overview with 
p <- ggpairs(human, lower = list(combo = wrap("facethist", bins = 20)))
p

```


```{r,fig.dim= c(8,6)}
#plotting the correlation matrix
cor_matrix <- cor(human) %>% round(digits = 2)

corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)

```


<br /> 

## 2 BPRS (8p)

<br /> *Implement the analyses of Chapter 9 of MABS using the BPRS data. (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)*

```{r,fig.dim= c(10,8)}
pca_human <- prcomp(human)

# making a summary of the PCA 
s <- summary(pca_human)

# printing rounded percentages of variance captured by each principal component
pca_pr <- round(100*s$importance[2, ], digits = 1)
pca_pr
# making the lables include the percentages
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1),
       col = c("cornsilk4", "darkgoldenrod"),
       xlab = pc_lab[1],
       ylab = pc_lab[2])

```




<br /> 

<br /> **THE END!**
<br />
