# Week 2: Regression and model validation

<br /> *Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

<br /> 

## 1 Description of the data

<br /> The data includes survey results from statistics course students in Finland regarding
their learning. Data is collected 2014-2015. After combining variables that measure the same dimension and excluding observations where the exam points are zero, the data consists of **166 observations and 7 variables** listed here: 

* gender
* age
* attitude
* deep approach (maximized understanding with a true commitment)
* surface approach (memorizing without understanding)
* strategic approach (applying strategy to maximize learning)
* points (max points)

The three learning approaches combine **8 subscales** that all include 4 questions:

* deep: Seeking meaning, relating ideas, use of evidence
* surface: lack of purpose, unrelated memorizing, syllabus-boundness
* strategic: organized studying, time management

Reading the data and checking the structure (working diary being IODS-project):

```{r}
data <- read.csv("data/learning2014", row.names = 1)
str(data)
```

<br /> 

## 2 Graphical overview of the data

<br /> Summary of the data --- basic statistics of the variables:

```{r}
summary(data)
```


Graphical overview using libraries GGally and ggplot2. The first variable, gender, is defining the colour and alpha sets transparency in the graphs so that the overlapping data can be seen:

```{r}
library(ggplot2)
library(GGally)
p <- ggpairs(data, mapping = aes(col=data$gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

The correlation coefficient, corr, refers to Pearson correlation by default --- correlations equal to +1 or −1 correspond to data points lying exactly on a line. Stars in the end the correlation value represent the p-values (as explained [here](https://ggobi.github.io/ggally/reference/ggally_cor.html)): 

* "***" p < 0.001
* "**" p < 0.01
* "*" p < 0.05
* "." p < 0.10

<br /> 

#### Interpretation:

There are almost twice as many female students than male. Majority of the students are between 20 and 30 years old. Male students have generally better attitude than female.

Points and attitude have the highest positive correlation 0.437, which is also significant (p < 0.001). Highest significant negative correlation can be seen between surface and deep learning approaches with corr = -0.324 (male up to -0.622). This is to be expected since they could be considered the opposite. 

Strategic learning approach is slightly more common among female students than male. 

Less significant negative correlation can be seen between surface and strategic learning approaches, corr = -0.161, and between surface learning approach and attitude, corr = -0.176.

The distribution of deep learning approach is shifted more towards higher values than with strategic or surface approach. The mean and median are approx. 3.6 --- for strategic approach approx 3.1 and surface approach 2.8.

To me it seems odd, that the points do not correlate with deep learning approach at all. **Could this mean that the exam/assignments do not measure the deep learning? Or that the deep learning approach does not in fact lead to learning?** There is more evidence (all though no significance shown) that the surface learning approach has some negative correlation and strategic approach positive with the points.


## 3 Multiple regression
<br /> Since the points have a highly significant correlation with attitude and a slightly significant correlation with surface and strategic learning approaches, I have chosen those three for the multiple regression as explanatory variables to the target variable "points". 

```{r}
my_model_multi <- lm(points ~ attitude + stra + surf, data = data)
summary(my_model_multi)
```

"comment and interpret the results. Explain and interpret the statistical test related to the model parameters. "

<br /> Since only the attitude has a high significance (***), I will try making models with attitude and surface approach, and attitude and strategic approach. In the latter model (shown below) the strategic learning approach has p-value rather close to 0.05, and therefore I will choose this model to further analyze.

```{r}
my_model_2 <- lm(points ~ attitude + stra, data = data)
summary(my_model_2)
```



