# Week 2: Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

## 1 Description of the data

The data includes survey results from statistics course students in Finland regarding
their learning. Data is collected 2014-2015. After combining variables that measure the same dimension and excluding observations where the exam points are zero, the data consists of **166 observations and 7 variables** listed here: 

* gender
* age
* attitude
* deep approach (maximized understanding with a true commitment)
* surface approach (memorizing without understanding)
* strategic approach (applying strategy to maximize learning)
* points (max points)

The three learning approaches combine **8 subscales** that all include 4 questions:

* deep: Seeking meaning, relating ideas, use of evidence
* surface: lack of purpose, unrelated memorizing, syllabus-boundness
* strategic: organized studying, time management

Reading the data and checking the structure (working diary being IODS-project):

```{r}
data <- read.csv("data/learning2014", row.names = 1)
str(data)
```


## 2 Graphical overview of the data

Summary of the data --- basic statistics of the variables:

```{r}
summary(data)
```


Graphical overview using libraries GGally and ggplot2. The first variable, gender, is defining the colour:

```{r}
library(ggplot2)
library(GGally)
p <- ggpairs(data, mapping = aes(col=data$gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

